---
title: "mLOX Slope Models"
output: html_document
date: "2025-12-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(tibble)
library(tidyr)


data <- read_excel("PLCO_longitudinal_mCAs_0s.xlsx") 

mlox <- data  %>%
  filter(type_FINAL == "mLOX")

```


# Models 

## Continuous Models 

### LME with log transformed data 

The Linear Mixed Model used is shown below:

Yij (Yij>0) = B0 + B1t + bi0 + e

```{r}
library(lme4)

mlox <- mlox %>%
  group_by(plco_id) %>%
  mutate(mca_num = dense_rank(mCA_code3)) %>%
  ungroup()

mlox <- mlox %>%
  mutate(cf_log = log(cf_FINAL))

mlox_pos <- mlox  %>%
  filter(cf_FINAL > 0)


lmm_log <- lmer(
  cf_log ~ STUDY_YR + 
    (1 + STUDY_YR | plco_id),
  data = mlox_pos
)

summary(lmm_log)

```


### Gamma Distribution

```{r}

library(glmmTMB)

glmm_gamma <- glmmTMB(
  cf_FINAL ~ STUDY_YR + 
    (1 + STUDY_YR | plco_id),
  family = Gamma(link = "log"),
  data = mlox_pos
)

summary(glmm_gamma)

```


## Logistic Models 

### GLMMIX

The Logistic Generalized Linear Mixed Model used is shown below:

logit(P(Yij>0)) = alpha0 + alpha1t + Ci0

```{r}
mlox <- mlox %>%
  mutate(
    cf_binary = ifelse(cf_FINAL > 0, 1, 0),
STUDY_YR = as.numeric(STUDY_YR)
)

glmmix <- glmer(cf_binary ~ STUDY_YR + (1 + STUDY_YR | plco_id), family = binomial(link = "probit"), data = mlox)

summary(glmmix)
```


## Joint Model 

### Baeysian

# log

```{r}
library(brms)

mlox$Ipos <- as.integer(mlox$cf_FINAL > 0)

log_brms <- brm(
  bf(
    Ipos ~ STUDY_YR + (1 + STUDY_YR | p | plco_id),
    family = bernoulli(link = "probit")
  ) +
  bf(
    cf_FINAL | subset(Ipos == 1) ~ STUDY_YR + (1 + STUDY_YR | p | plco_id),
    family = lognormal()
  ),
  data = mlox,
  backend = "cmdstanr",
  threads = threading(4),
  chains = 4,
  cores = 4,
  iter = 1500,
  warmup = 750,
  sample_prior = "no",
  control = list(
    adapt_delta = 0.85,
    max_treedepth = 12
  )
)

print(summary(log_brms), digits = 3)

```



# gamma 

```{r}

library(brms)

gamma_brms <- brm(
 bf(Ipos ~ STUDY_YR + (1 + STUDY_YR | p | plco_id),
     family = bernoulli(link = "probit")) +
    bf(cf_FINAL | subset(Ipos == 1) ~ STUDY_YR + (1 + STUDY_YR  | p | plco_id),
       family = Gamma(link = "log")),
  data = mlox,
  backend = "cmdstanr",           
  threads = threading(4),         
  chains = 4,
  cores = 4,
  iter = 1500,                    
  warmup = 750,
  sample_prior = "no",            
  control = list(
    adapt_delta = 0.85,           
    max_treedepth = 12   
  
)
)

print(summary(gamma_brms), digits = 3)

```

## Comparing joint model and seperate model estimates for gamma distribution 

```{r}

results <- data.frame(
  Term = c(
    "B0", "B1",
    "alpha0", "alpha1",
    
    "sigma(b0)", "sigma(b1)",
    "sigma(c0)", "sigma(c1)",
    "cor(b0,b1)", "cor(b0,c0)", "cor(b1,c0)",
    "cor(b0,c1)", "cor(b1,c1)", "cor(c0,c1)"
  ),
  
  Combined_Estimate = c(
    -3.195,  0.047,
    1.323,  0.051,
    
     1.035,  0.090,
     1.087,  0.069,
     0.052,  0.825,  0.264,
     0.033,  0.128, -0.173
  ),
  
  Combined_SE = c(
    0.080, 0.009,
    0.182, 0.046,
    
    0.194, 0.059,
    0.060, 0.018,
    0.425, 0.096, 0.389,
    0.240, 0.395, 0.147
  ),
  
  Separate_Estimate = c(
    -3.184517, 0.047768,
     3.87888, -0.10703,
    
     1.07609, 0.07134,
     5.2157,  0.9789,
    
    -0.83, NA, NA,
     NA, NA, -0.19
  ),
  
  Separate_SE = c(
    0.078324, 0.008715,
    0.30288, 0.08686,
    
    NA, NA,
    NA, NA,
    
    NA, NA, NA,
    NA, NA, NA
  )
)

print(results)

```




## Comparing joint model and seperate model estimates for log tranformed normal distribution 

```{r}

results <- data.frame(
  Term = c(
    "B0", "B1",
    "alpha0", "alpha1",
    
    "sigma(b0)", "sigma(b1)",
    "sigma(c0)", "sigma(c1)",
    "cor(b0,b1)", "cor(b0,c0)", "cor(b1,c0)",
    "cor(b0,c1)", "cor(b1,c1)", "cor(c0,c1)"
  ),
  
  Combined_Estimate = c(
    -3.238,  0.047,
     1.319,  0.052,
    
     1.032,  0.091,
     1.100,  0.085,
     0.051,  0.808,  0.278,
     0.081,  0.064, -0.221
  ),
  
  Combined_SE = c(
    0.083, 0.010,
    0.185, 0.046,
    
    0.198, 0.058,
    0.061, 0.013,
    0.420, 0.099, 0.381,
    0.231, 0.435, 0.115
  ),
  
  Separate_Estimate = c(
    -3.223027, 0.048199,
     3.87888, -0.10703,
    
     1.0929, 0.0847,
     5.2157, 0.9789,
    
    -0.24, NA, NA,
     NA, NA, -0.19
  ),
  
  Separate_SE = c(
    0.079378, 0.009304,
    0.30288, 0.08686,
    
    NA, NA,
    NA, NA,
    
    NA, NA, NA,
    NA, NA, NA
  )
)

print(results)


```



## Tobit Model 

```{r}

censor_point <- min(mlox$cf_log[is.finite(mlox$cf_log)])
censor_point

mlox$cf_log[!is.finite(mlox$cf_log)] <- censor_point

mlox$cens_type <- ifelse(mlox$cf_log == censor_point, -1, 0)

tobit_brms <- brm(
  bf(cf_log | cens(cens_type) ~ STUDY_YR + (1 + STUDY_YR | ID | plco_id)),
  family = gaussian(),
  data = mlox,
  chains = 4, cores = 4
)

summary(tobit_brms)
```


```{r}

results <- data.frame(
  Term = c("Intercept", "STUDY_YR", "sigma(b0)","sigma(b1)", "p(rho)", "sigma(e)"),
  
  Estimate = c(-3.53, 0.10, 1.26, 0.05, 0.01, 0.72),
  SE       = c(0.06, 0.02, 0.08, 0.03, 0.42, 0.03)
)

print(results)

```

#########################################################################################################

## Becky's analysis 

### Becky's Model 


### LME 

```{r}
library(dplyr)
library(lme4)

mlox <- mlox %>%
  mutate(
    age_65 = age_at_collection - 65,
    smoking_status = factor(smoking_status, levels = c("Never", "Former", "Current"))
  )

mlox_pos <- mlox  %>%
  filter(cf_FINAL > 0)

mlox_lme <- lmer(
  cf_FINAL ~ age_65 * smoking_status + 
    (1 + age_65 | plco_id),
  data = mlox_pos,
  REML = FALSE
)

summary(mlox_lme)
```
### GLMMIX

```{r}

mlox <- mlox %>%
  mutate(
    cf_binary = ifelse(cf_FINAL > 0, 1, 0),
    age_65 = age_at_collection - 65,
    smoking_status = factor(smoking_status, levels = c("Never", "Former", "Current"))
  )

mlox_glmmix <- glmer(
  cf_binary ~ age_65 * smoking_status + 
    (1 + age_65 | plco_id),
  family = binomial(link = "probit"),
  data = mlox
)

summary(mlox_glmmix)
```

### hurdle model 

```{r}

library(dplyr)
library(brms)

mlox <- mlox %>%
  mutate(
    Ipos = as.integer(cf_FINAL > 0),
    cf_pos = ifelse(cf_FINAL > 0, cf_FINAL, NA_real_),
    age_65 = age_at_collection - 65,
    smoking_status = factor(smoking_status, levels = c("Never", "Former", "Current"))
  )

mlox_hurdle <- brm(
  bf(
    Ipos ~ age_65 * smoking_status + (1 + age_65 | p | plco_id),
    family = bernoulli(link = "probit")
  ) +
  bf(
    cf_pos | subset(Ipos == 1) ~ age_65 * smoking_status + (1 + age_65 | p | plco_id),
    family = gaussian()
  ),
  data = mlox,
  backend = "cmdstanr",
  threads = threading(4),
  chains = 4,
  cores = 4,
  iter = 2000,
  warmup = 750,
  sample_prior = "no",
  control = list(
    adapt_delta = 0.99,
    max_treedepth = 12
  )
)

print(summary(mlox_hurdle), digits = 3)
```


### Table 

```{r}
results <- data.frame(
  Term = c(
    "B0", "B1", 
    "B2_Former", "B2_Current",
    "B3_Former", "B3_Current",
    "alpha0", "alpha1",
    "alpha2_Former", "alpha2_Current",
    "alpha3_Former", "alpha3_Current",
    "sigma(b0)", "sigma(b1)",
    "sigma(c0)", "sigma(c1)",
    "cor(b0,b1)", "cor(b0,c0)", "cor(b1,c0)",
    "cor(b0,c1)", "cor(b1,c1)", "cor(c0,c1)", "sigma(e)"
  ),
  
  Combined_Estimate = c(
    0.065, 0.003,
    0.065, 0.038,
    0.002, -0.002,
    1.287, 0.089,
    0.619, 0.097,
   -0.037, 0.001,
    1.247, 0.097,
    0.162, 0.004,
    0.282, 0.711, 0.631,
    0.069, 0.492, 0.344,
    0.036
  ),
  
  Combined_SE = c(
    0.015, 0.001,
    0.025, 0.037,
    0.002, 0.003,
    0.195, 0.030,
    0.321, 0.391,
    0.043, 0.057,
    0.235, 0.041,
    0.008, 0.002,
    0.309, 0.123, 0.256,
    0.330, 0.318, 0.232,
    0.002
  ),
  
  Separate_Estimate = c(
    0.068, 0.003,
    0.065, 0.038,
    0.002, -0.003,
    1.844, 0.167,
    0.400, -0.027,
   -0.070, 0.012,
    1.485, 0.099,
    0.160, 0.004,
    1.000, NA, NA,
    NA, NA, 0.410,
    0.036
  ),
  
  Separate_SE = c(
    0.015, 0.001,
    0.025, 0.035,
    0.002, 0.003,
    0.369, 0.042,
    0.287, 0.395,
    0.043, 0.060,
    NA, NA,
    NA, NA,
    NA, NA, NA,
    NA, NA, NA,
    NA
  )
)

print(results)

```



### Tobit Model 

```{r}
library(brms)

censor_point <- min(mlox$cf_FINAL[is.finite(mlox$cf_FINAL)])

mlox$cf_FINAL[!is.finite(mlox$cf_FINAL)] <- censor_point

mlox$cens_type <- ifelse(mlox$cf_FINAL == censor_point, -1, 0)

mlox_tobit <- brm(
  bf(
    cf_FINAL | cens(cens_type) ~ age_65 * smoking_status + (1 + age_65 | p | plco_id)
  ),
  family = gaussian(),
  data = mlox,
  chains = 4,
  cores = 4
)

print(summary(mlox_tobit), digits = 3)
```

```{r}
results <- data.frame(
  Term = c(
    "Intercept", 
    "age_65",
    "smoking_Former",
    "smoking_Current",
    "age_65:smoking_Former",
    "age_65:smoking_Current",
    "sigma(b0)", 
    "sigma(b1)", 
    "p(rho)", 
    "sigma(e)"
  ),
  
  Estimate = c(
    0.049,
    0.004,
    0.067,
    0.037,
   -0.003,
   -0.002,
    0.160,
    0.002,
    0.179,
    0.057
  ),
  
  SE = c(
    0.015,
    0.001,
    0.025,
    0.036,
    0.002,
    0.003,
    0.008,
    0.001,
    0.459,
    0.002
  )
)

print(results)
```



## Figure 3 

### LME Plot 

```{r}

library(lme4)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggeffects)

age_fix <- function(x) {
  x + 65
}

# Create plot
plot_curve <- function(lmer_model, predict_term, graph_name) {
  
  Profile <- predict_response(lmer_model, terms = predict_term)
  
  plot(Profile) +
    labs(x = "Age (Years)", y = "Clonal Fraction") +
    scale_color_manual(
      values = c(
        "Never" = "#D55E00",
        "Current" = "#0072B2",
        "Former" = "#009E73"
      )
    ) +
    scale_fill_manual(
      values = c(
        "Never" = "#D55E00",
        "Current" = "#0072B2",
        "Former" = "#009E73"
      )
    ) +
    scale_x_continuous(label = age_fix, limits = c(-11, 14),
                       breaks = c(-10, -5, 0, 5, 10, 14)) +
    ggtitle(graph_name) +
    theme_minimal() +
    theme(legend.position = "none")
}


mLOX_fit <- lmer(cf_FINAL ~ age_65 + 
                  smoking_status + 
                  smoking_status*age_65 +
                  (age_65|plco_id), 
                 data=mlox)

summary(mLOX_fit)


#Create Graph for results
p_lme <- plot_lmer_curve(mLOX_fit, 
                      predict_term = c("age_65","smoking_status"),
                      "mLOY") +
  labs(title = NULL, x = "Age (Years)", y = "Clonal Fraction", smoking_status = "Smoking Status") +
  scale_x_continuous(label = age_fix, limits = c(-11,14), breaks = c(-10,-5,0,5,10,14)) +
  scale_y_continuous(limits = c(-0.10, 0.60), breaks = c(0.00,0.10,0.20,0.30,0.40,0.50,0.60)) +
  ggtitle("mLOX LMM") +
  theme_minimal() + 
    theme(
  legend.position = "none",
  axis.title.x = element_text(size = 18),
  axis.title.y = element_text(size = 18),
  axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14),
  plot.title = element_text(size = 20, face = "bold"),
  legend.title = element_text(size = 16),
  legend.text = element_text(size = 14)
)

#Inspect Graph
p_lme

ggsave("mLOX_LMM.pdf", p_lme, width = 6, height = 5)

```

### Hurdle Plot 

```{r}

library(dplyr)
library(tidyr)
library(ggplot2)

age_fix <- function(x) {
  x + 65
}

# Prediction grid
pred_grid <- expand_grid(
  age_65 = seq(min(mlox$age_65, na.rm = TRUE),
               max(mlox$age_65, na.rm = TRUE),
               length.out = 500),
  smoking_status = levels(mlox$smoking_status),
  Ipos = 1L
)

# Get predictions for positive part
p_cf_df <- fitted(
  mlox_hurdle,
  newdata = pred_grid,
  resp = "cfpos",
  re_formula = NA,
  summary = TRUE
) %>%
  as.data.frame() %>%
  bind_cols(pred_grid)

p_cf_df$smoking_status <- factor(
  p_cf_df$smoking_status,
  levels = c("Never", "Former", "Current")
)

p_hurdle <- ggplot(p_cf_df, aes(x = age_65, y = Estimate, color = smoking_status, fill = smoking_status)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.10, color = NA) +
  geom_line(linewidth = 1) +
  labs(title = NULL, x = "Age (Years)", y = "Clonal Fraction") +
  scale_color_manual(
    values = c(
      "Never" = "#D55E00",
      "Current" = "#0072B2",
      "Former" = "#009E73"
    )
  ) +
  scale_fill_manual(
    values = c(
      "Never" = "#D55E00",
      "Current" = "#0072B2",
      "Former" = "#009E73"
    )
  ) +
  scale_x_continuous(
    labels = age_fix,
    limits = c(-11, 14),
    breaks = c(-10, -5, 0, 5, 10, 14)
  ) +
  scale_y_continuous(
    limits = c(-0.10, 0.60),
    breaks = c(0.00, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60)
  ) +
  ggtitle("mLOX Hurdle") +
  theme_minimal() +
    theme(
  legend.position = "none",
  axis.title.x = element_text(size = 18),
  axis.title.y = element_text(size = 18),
  axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14),
  plot.title = element_text(size = 20, face = "bold"),
  legend.title = element_text(size = 16),
  legend.text = element_text(size = 14)
)

# Show plot
p_hurdle

ggsave("mLOX_hurdle.pdf", p_hurdle, width = 6, height = 5)

```



## Tobit Plot 

```{r}
library(brms)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggeffects)

p_tobit <- plot_curve(
  mlox_tobit,
  predict_term = c("age_65", "smoking_status"),
  "mLOX"
) +
  labs(
    title = NULL,
    x = "Age (Years)",
    y = "Clonal Fraction",
    smoking_status = "Smoking Status"
  ) +
  scale_x_continuous(
    labels = age_fix,
    limits = c(-11, 14),
    breaks = c(-10, -5, 0, 5, 10, 14)
  ) +
  scale_y_continuous(
    limits = c(-0.10, 0.60),
    breaks = c(0.00, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60)
  ) +
  ggtitle("mLOX Tobit") +
  theme_minimal() +
  theme(
  legend.position = "none",
  axis.title.x = element_text(size = 18),
  axis.title.y = element_text(size = 18),
  axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14),
  plot.title = element_text(size = 20, face = "bold"),
  legend.title = element_text(size = 16),
  legend.text = element_text(size = 14)
)

p_tobit

ggsave("mLOX_tobit.pdf", p_tobit, width = 6, height = 5)
```

```{r}

mLOX_plot <- p_lme + p_hurdle + p_tobit +
  plot_layout(ncol = 3)

mLOX_plot

ggsave("mLOX_plots.png", mLOX_plot, width = 14, height = 4)

```